---
title: "ESM 244 Lab 1"
author: "Renee LaManna"
date: "1/9/2022"
output: html_document
---

```{r setup, include=TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(here)
library(sf)
library(tmap)

### install.packages('tmap')
### update.packages(ask = FALSE)
```

### Read in the data

```{r}
sf_trees <- read_csv(here("data", "sf_trees", "sf_trees.csv"), show_col_types = FALSE)
# names(sf_trees) - will give us column names
# summary(sf_trees) - will give us summary 

```

### Part 1: wrangling and ggplot review

**Example 1:** Find counts of observation by `legal_status` & wrangle a bit

```{r}
### method 1: group_by() %>%  summarize ()
sf_trees %>% 
  group_by(legal_status) %>% 
  summarize(tree_count = n()) # Number of trees will be the number of rows in our data frame

### method 2: different way plus a few new fxns
top_5_status <- sf_trees %>% 
  count(legal_status) %>% 
  drop_na(legal_status) %>% 
  rename(tree_count = n) %>% 
  relocate(tree_count) %>% 
  slice_max(tree_count, n = 5) %>% 
  arrange(-tree_count) # can also use desc(tree_count) fxn
  
  
```

Make a graph of the top 5 from above

``` {r}
ggplot(data = top_5_status, aes(x = fct_reorder( legal_status, tree_count), y = tree_count)) +
  geom_col(fill = 'seagreen4') +
  labs(x = 'Legal status', y = 'Tree count') +
  coord_flip() +
  theme_minimal()
       

```

**Example 2:** Only going to keep observations where legal status is "Permitted Site" and caretaker is "MTA", and store as 'permitted_data_df' 

shift-cmd-c to comment/uncomment active/nonactive quickly
```{r}
# sf_tree$legal_status %>%  unique() # take legal status column out and get rid of any duplicates
# unique(sf_trees$caretaker)
permitted_data_df <- sf_trees %>% 
  filter(legal_status %in% c('Permitted Site', 'Private') & caretaker %in% 'MTA')

# both conditions must be meant for it to keep the line, can also use "&" instead of ","
# If u want to read it as or, use "|"


```

**Example 3:** Only keep Blackwood Acacia trees, and then only keep columns `legal_status`, `date`, `latitude`, `longitude` and store as `blackwood_acacia_df`

```{r}
blackwood_acacia_df <- sf_trees %>% 
  filter(str_detect(species, 'Blackwood Acacia')) %>% 
  select(legal_status, date, lat = latitude, lon = longitude)

### Make a little graph of locations
ggplot(data = blackwood_acacia_df, aes(x = lon, y =lat)) +
  geom_point(color = 'darkgreen')
```

**Example 4:** use `tidyr::separate()`

```{r}
sf_trees_sep <- sf_trees %>% 
  separate(species, into = c('spp_scientific', 'spp_common' ), sep = ' :: ')
```

**Example 5:** use `tidyr::unite()`

```{r}
ex_5 <- sf_trees %>% 
  unite('id_status', tree_id, legal_status, sep = '_COOL_')
```

